### 2023-06-08 paper/localpsf.tex
Update abstract to include Noemi's edits, plus some of my additional edits

% REQUIRED
\begin{abstract}
	We present an efficient matrix-free point spread function (PSF) method for approximating operators that have locally supported non-negative integral kernels.
	The method computes impulse responses of the operator at a collection of scattered points, and interpolates translated and scaled versions of these impulse responses to approximate entries of the integral kernel. By applying the operator to Dirac combs associated with batches of point sources, we compute a large number of impulse responses using a small number of operator applies. We solve an ellipsoid packing problem to select as many point sources as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. Support ellipsoids are estimated a-priori based on impulse response moments, which are computed by applying the operator to a small number of polynomial functions. The interpolation uses a local mean displacement invariance approximation, which generalizes classical local translation invariance. The ability to rapidly evaluate kernel entries allows us to construct a hierarchical matrix (H-matrix) approximation of the operator. Further matrix computations are performed with H-matrix methods. We illustrate the method by approximating the Hessians in two example problems: an ice sheet flow inverse problem governed by a Stokes partial differential equation (PDE), and a diffusive transport inverse problem governed by an advection-diffusion PDE. Because the data in these inverse problems are highly informative about the unknown parameters, the data misfit terms in the Hessians are high rank. This makes conventional matrix-free Hessian approximation methods, which are based on low rank approximation, extremely expensive in terms of the required number of Hessian operator applies. In contrast, the numerical results show that the proposed method can form good approximations of these high rank Hessians using a small number of Hessian operator applies. We use the Hessian approximations as preconditioners to accelerate the solution of the inverse problems. Numerical results demonstrate that this approach substantially reduces the number of PDE solves required to solve the inverse problems.

>>>>

% REQUIRED
\begin{abstract}
	We present an efficient matrix-free point spread function (PSF) method for approximating operators that have locally supported non-negative integral kernels. The method computes impulse responses of the operator at a collection of scattered points, and interpolates translated and scaled versions of these impulse responses to approximate entries of the integral kernel. By applying the operator to Dirac combs associated with batches of point sources, we compute a large number of impulse responses using a small number of operator applies. We solve an ellipsoid packing problem to select as many point sources as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. Support ellipsoids are estimated a-priori based on impulse response moments, which are computed by applying the operator to a small number of polynomial functions. The interpolation uses a local mean displacement invariance approximation, which generalizes classical local translation invariance. The ability to rapidly evaluate kernel entries allows us to construct a hierarchical matrix (H-matrix) approximation of the operator. Further matrix computations are performed with H-matrix methods. We apply the proposed method to approximate the Hessian operator arising in inverse problems governed by partial differential equations (PDEs). While for many ill-posed inverse problems the Hessian of the data misfit term exhibits a low rank structure, and hence a low rank approximation is suitable, for many problems of practical interest the numerical rank of this operator is still large. This makes conventional matrix-free Hessian approximation methods expensive in terms of the required number of Hessian operator applies (hence PDE solves). In contrast, the performance of the proposed method depends on the locality and shape of the impulse response rather than the numerical rank. Hessian impulse responses typically become more local as the numerical rank increases, which makes this operator a perfect candidate for the proposed approach. We illustrate the proposed approach with two specific problems: inversion for the basal sliding coefficient in an ice sheet flow problem and inversion for the initial condition in a diffusive transport problem. To demonstrate the effectiveness of the approximation, we use the proposed approach to build preconditioners to accelerate the solution of the inverse problems. We compare the results with classical regularization-based preconditioning and no preconditioning. The numerical results demonstrate the superior performance (measured in number of PDE solves) of the inverse solver when using the proposed PSF-based preconditioner. Finally, we present a comprehensive numerical study for the influence of the diffusion parameters and simulation time (for the diffusive transport problem) on the effectiveness of the approximation. These parameters have an effect on the shape of the impulse responses and on the rank of the Hessian. The results reveal that PSF-based preconditioners are able to form good approximations of high-rank Hessians using a small number of Hessian operator applies.

%	We present an efficient matrix-free point spread function (PSF) method for approximating operators that have locally supported non-negative integral kernels.
%	The method computes impulse responses of the operator at a collection of scattered points, and interpolates translated and scaled versions of these impulse responses to approximate entries of the integral kernel. By applying the operator to Dirac combs associated with batches of point sources, we compute a large number of impulse responses using a small number of operator applies. We solve an ellipsoid packing problem to select as many point sources as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. Support ellipsoids are estimated a-priori based on impulse response moments, which are computed by applying the operator to a small number of polynomial functions. The interpolation uses a local mean displacement invariance approximation, which generalizes classical local translation invariance. The ability to rapidly evaluate kernel entries allows us to construct a hierarchical matrix (H-matrix) approximation of the operator. Further matrix computations are performed with H-matrix methods. We illustrate the method by approximating the Hessians in two example problems: an ice sheet flow inverse problem governed by a Stokes partial differential equation (PDE), and a diffusive transport inverse problem governed by an advection-diffusion PDE. Because the data in these inverse problems are highly informative about the unknown parameters, the data misfit terms in the Hessians are high rank. This makes conventional matrix-free Hessian approximation methods, which are based on low rank approximation, extremely expensive in terms of the required number of Hessian operator applies. In contrast, the numerical results show that the proposed method can form good approximations of these high rank Hessians using a small number of Hessian operator applies. We use the Hessian approximations as preconditioners to accelerate the solution of the inverse problems. Numerical results demonstrate that this approach substantially reduces the number of PDE solves required to solve the inverse problems.
